---
title: "Data Analysis"
output: html_document
---

# Introduction

We are going to be doing analysis on the Iris data set.
As mentioned before, you can do analysis on your own data set, but for this example
we will be doing it on a built in data set.

The head function will allow us to get an idea of how our dataset is structured
```{r}
head(iris)
```

## Creating a data object

While doing analysis it is useful to be able to break up our data into smaller
data objects. We can use the code below to do so. I will explain what this code is doing later.
```{r}
setosa <- iris[iris$Species == setosa, ]
```

When running this line of code we see we get an error, this is because when
checking the conditional statement `iris$Species == setosa` R is trying to compare
`iris$Species` to `setosa` the data object.

We can fix this issue by putting `setosa` in quotation marks, which tells R that this is a word.
```{r}
setosa <- iris[iris$Species == "setosa",]
virginica <- iris[iris$Species == "virginica",]
versicolor <- iris[iris$Species == "versicolor",]
```

**Now lets break down what exactly is going on in each one of these lines.**
`setosa` is the variable of the data object we are creating
`<-` says that whatever is happening on the right will be assigned to the variable
`iris[]` is the data set which we are pulling the information from
`iris$Species` is the column from the iris data set which we are going to be comparing too.
`==` evaluates to true if both sides are the same.
`"setosa"` is the word which we are going to be comparing too.
`,` defines that this is the end of a line within the data object we are creating.

Think about the format of a CSV file which our data sets are, they look something like:
---
sepal_length,sepal_width,petal_length,petal_width,species
5.1,3.5,1.4,0.2,setosa
4.9,3.0,1.4,0.2,setosa
7.0,3.2,4.7,1.4,versicolor
6.4,3.2,4.5,1.5,versicolor
6.3,3.3,6.0,2.5,virginica
5.8,2.7,5.1,1.9,virginica
---

When one of the above lines of code is ran it goes through every line of the dataset, and compares
the species to what is defined in the code. If the two are a match, it adds it to the new data object
otherwise it just moves on to the next line.

## Calculating mean, and median.

The mean is important in many statistical calculations. Lets calculate it the hard way, then the easy way. This will demonstrate how some stat functions in R simplify calculations.
```{r}
mean <- sum(setosa$Petal.Length) / length(setosa$Petal.Length)
print(mean)
```
This is the hard way, and I say that very lightly. We calculated the sum of the entries in the column, and then divided it by the length of the column (how many entries are in it).

Now the easy way:
```{r}
mean <- mean(setosa$Petal.Length)
print(mean)
```
This did all the calculations for us.

**Challenge**
Try to calculate the mean and median for a different entire in any of the three data objects.
Use the function `meadian()` and `mean()`

```{r}
mean <- # mean()
print(mean)
median <- # median()
print(median)
```

More useful statistical functions can be found [here](https://www.statmethods.net/management/functions.html)


## Doing t-tests

The purpose of a t-test is to determine if there is a significant difference between the mean of two
different groups which are related and if two data sets come from the same population.

Knowing what t-tests are now, we can perform one using the function `t.test(x = val1, x = val2)`
```{r}
ttest_result <- t.test(x = setosa$Petal.Length, y = versicolor$Petal.Length)
ttest_result
```

Because our calculated p-value is less then our alpha, we are going to reject our null hypothesis that these two species have the same petal length.
Our alpha is calculated by doing 1 - confidence_interval (e.g. 1 - 0.95).

**Now lets break down what exactly is going on in each one of these lines.**
`ttest_result` is the variable which we are storing the t-test in
`<-` says that whatever is happening on the right will be assigned to the variable
`t.test()` is our function
`x = setosa$Petal.Length` is our x values for the t-test
`y = setosa$Petal.Length` is our y values for the t-test

What is different between displaying this calculation and the mean for example is the need for a
print statement. We displayed the output of the t-test by just calling the variable we assigned it too.

# Doing ANOVA tests

An ANOVA test does exactly what it stands for: Analysis of Variance. An ANOVA test allows us to determine if there is a difference between groups within a population.

```{r}
anova_result <- aov(formula = Petal.Length ~ Species, data = iris)
summary(object = anova_result)
```

**Now lets break down what exactly is going on in each one of these lines.**
`anova_result` is the variable which we are storing the t-test in
`<-` says that whatever is happening on the right will be assigned to the variable
`aov()` is our function
` forumula = Petal.Length ~ Species` is our function for the test, we are going to be analyzing the variance of petal length on the species
`data = iris` is the data set which the information is coming from.

To see the result of our ANOVA test we called the `summary()` function. This is different then the t-test and mean/median as done before.

# Basic graphing functions

Plotting is very important in statistics, and with R we can make graphs easily.
These will be the most basic forms of plots, the next section will cover plotting more in depth.


Here is a basic plot.
```{r}
x <- seq(-pi, pi, 0.05)
plot(x, sin(x))
```
`x` contain values from -pi to pi by 0.05 increments
`plot()` is the plotting function
Within `plot()`: `x` is our x values, and `sin(x)` is our function.


Here is a basic box plot.
```{r}
boxplot(iris$Petal.Length ~ iris$Species)
```
`boxplot()` is the plotting function
`iris$Petal.Length ~ iris$Species` is the function comparing petal length and species.

```{r}
hist(iris$Petal.Length)
```
`hist()` is the plotting function
`iris$Petal.Length` is the data we are plotting



